\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-29 tex/paper.tex
save

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we use a univariate Gaussian proposal distribution
$q(\log \lambda|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             (Top left) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution with, in this case, qualitatively
                        better results for the penalized model.
             (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $p(\theta|\psi)$ after adversarial variational optimization.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\theta|\psi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with qualitatively  better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the
non-differentiability of the underlying generator, this confirms that inference
with adversarial variational optimization works.

@@

\glnote{use 5d example from carl.}
\glnote{contour plots for $\alpha$ and $\beta$.}

>>>>

The observed data is sampled from a Poisson with scale $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 300 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.01$, decay rates
$\beta_1=\beta_2=0.5$ and its inner first and second moment vectors reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we parameterize $\theta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\theta|\psi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\psi$ are updated by taking one Adam step, with
$\alpha=0.01$ and $\beta_1=\beta_2=0.1$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             ({\it Top left}) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             ({\it Top right}) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution.
             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $p(\theta|\psi)$ after adversarial variational optimization.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\theta|\psi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with in this case slightly  better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the
non-differentiability of the underlying generator, this confirms that inference
with adversarial variational optimization works.

@@

As a second example, we evaluate parameter inference for a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\mathbf{z}$ into the observed space.
\end{itemize}
We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored Gaussian
proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
initialized with zero mean and unit variance. The entropy penalty is set
to $\gamma=5$, while all other hyper-parameters are otherwise the same as in the previous example.

Starting with a proposal distribution $q(\theta|\psi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
inference quickly converges towards a proposal distribution whose density
concentrates around the nominal values $\theta^*$, as shown in the right plot of Figure~\ref{fig:multi}.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Likelihood $q(\theta|\psi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Likelihood $q(\theta|\psi)$ after adverarial variational optimization. The proposal density correctly concentrates around $\theta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}
